---
title: "SEAMAAC Heat Vulnerability Index"
subtitle: "An Expanded Small-Area HVI for South Philadelphia"
author: "Prepared for SEAMAAC by the Weitzman School of Design's Spring 2026 Planning HCED Practicum (and Claude)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    code_folding: show
    enclosed: true
    css: "styles.css"
---

```{=html}
<style>
  @import url('https://fonts.googleapis.com/css2?family=Exo+2:ital,wght@0,400;0,600;0,700&family=Open+Sans:wght@400;600;700&display=swap');
  body, p, li, td, th { font-family: 'Open Sans', sans-serif; font-size: 15px; color: #2b2b2b; }
  h1, h2, h3, h4, h5, h6 { font-family: 'Exo 2', sans-serif; font-weight: 700; color: #1f3b57; }
  h1.title { font-size: 30px; }
  .callout {
    background-color: #f5f7fa;
    border-left: 5px solid #4a7ca0;
    padding: 12px 16px;
    margin: 14px 0;
    border-radius: 4px;
    font-size: 14px;
  }
  .callout-title { font-family: 'Exo 2', sans-serif; font-weight: 700; color: #1f3b57; margin-bottom: 4px; }
  table { font-family: 'Open Sans', sans-serif; }
  caption { font-family: 'Exo 2', sans-serif; font-weight: 600; color: #1f3b57; }
</style>
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 7,
  dpi     = 150
)

options(scipen = 999)
```

# About this document

This report builds a **Heat Vulnerability Index (HVI)** for a small group of census tracts in South Philadelphia served by SEAMAAC. It extends the existing city-wide Philadelphia HVI by adding language-access and housing-quality indicators that are especially relevant to immigrant and refugee communities.

Every number and map in this document is generated directly from the code below — nothing is pasted in by hand. The index is fully reproducible: a collaborator can clone the file, change the study area or variable list in the **Configuration** chunk at the top, and re-knit the document to produce a new index with all tables and maps refreshed automatically.

:::: callout
::: callout-title
What is an HVI?
:::

A Heat Vulnerability Index is a score we give to a neighborhood that tells us how much harm residents are likely to experience during extreme heat. Scores are built from things we can measure — like poverty rates, housing quality, or how many people speak limited English. Higher scores mean more vulnerability.
::::

------------------------------------------------------------------------

# 1. Configuration

Everything you can change about the analysis lives here. **Do not hand-edit anything downstream.**

```{r config}
# ---------------------------------------------------------------------------
# Study area
# ---------------------------------------------------------------------------
study_area <- list(
  state       = "PA",
  county      = "Philadelphia County",
  acs_year    = 2024,
  acs_survey  = "acs5",
  # Last 6 digits of GEOID for each tract in the study area
  tract_codes = c("004001", "004002", "004101", "004103",
                  "004104", "004201", "004202", "037200")
)

# ---------------------------------------------------------------------------
# Color palette (used for every map and accent element)
# Change any hex code here and it propagates everywhere.
# ---------------------------------------------------------------------------
pal <- list(
  navy       = "#1f3b57",  # headings, strong accents
  steel      = "#4a7ca0",  # primary accent
  sage       = "#7aa88c",  # secondary accent, low-vulnerability fill
  sand       = "#e7d9b8",  # neutral background, mid-scale
  coral      = "#d97757",  # warning / mid-high vulnerability
  rust       = "#a83c2b"   # highest vulnerability
)

# Ordered gradient for choropleth scales (low -> high vulnerability)
hvi_gradient <- c(pal$sage, pal$sand, pal$coral, pal$rust)

# ---------------------------------------------------------------------------
# Index weighting — how much each sub-index contributes to the final HVI
# Housing & Economic PCA and Language Access PCA combine using these weights.
# Weights are normalized internally, so any positive numbers will work.
# ---------------------------------------------------------------------------
subindex_weights <- c(housing_econ = 0.5, language = 0.5)

# ---------------------------------------------------------------------------
# Variable specification
#
# Each entry defines ONE variable used in the index.
#
# Columns:
#   friendly_name   — the name used in code and tables
#   acs_numerator   — ACS variable code for the numerator (or the variable
#                     itself if it's already a rate / median)
#   acs_denominator — ACS variable code for the denominator. Use NA if the
#                     variable is already a rate, percentage, or median.
#   transform       — "none", "log", or "rate" ("rate" means num/denom)
#   direction       — +1 if HIGHER value means MORE vulnerable
#                     -1 if HIGHER value means LESS vulnerable
#   domain          — "housing_econ" or "language"
#   label           — human-readable label for tables
#
# To ADD a variable: append a row.
# To REMOVE a variable: delete or comment out its row.
# Nothing else needs to change.
# ---------------------------------------------------------------------------
var_spec <- tibble::tribble(
  ~friendly_name,            ~acs_numerator,                              ~acs_denominator, ~transform, ~direction, ~domain,        ~label,
  # --- Housing & Economic (8 core indicators) ---
  "med_value",               "B25107_001",                                NA_character_,    "none",      -1,        "housing_econ", "Median home value ($)",
  "med_hh_income",           "B19013_001",                                NA_character_,    "log",       -1,        "housing_econ", "Median household income (log $)",
  "pct_severely_burdened",   "B25070_010",                                "B25070_001",     "rate",      +1,        "housing_econ", "% renters paying 50%+ of income on rent",
  "pct_owner_occupied",      "B25003_002",                                "B25003_001",     "rate",      -1,        "housing_econ", "% owner-occupied households",
  "pct_poverty_below",       "B16009_002",                                "B16009_001",     "rate",      +1,        "housing_econ", "% below poverty (age 5+)",
  "pct_overcrowded",         "B25014_005,B25014_006,B25014_011,B25014_012","B25014_001",    "rate",      +1,        "housing_econ", "% households with >1 occupant per room",
  "pct_substandard_housing", "B25047_003,B25051_003",                     "B25002_001",     "rate",      +1,        "housing_econ", "% units lacking complete plumbing or kitchen",
  "pct_recent_movers",       "B25038_003,B25038_010",                     "B25038_001",     "rate",      +1,        "housing_econ", "% households moved in within last 4 years",
  # --- Language / Social (2 indicators; LEP added automatically below) ---
  "pct_ling_isolated",       "B16002_004,B16002_007,B16002_010,B16002_013","B16002_001",    "rate",      +1,        "language",     "% linguistically isolated households"
)

# LEP (B16004) — the table has many cells. We identify them programmatically
# below so we don't have to list each one by hand.
use_lep <- TRUE
```

:::: callout
::: callout-title
How the spec works
:::

The `var_spec` table is the single source of truth. Each row describes one variable: where to get it from the Census (`acs_numerator`, `acs_denominator`), how to transform it (`rate`, `log`, or `none`), and — most importantly — whether a higher value means a neighborhood is **more** or **less** vulnerable (`direction`). Everything after this is automatic.
::::

------------------------------------------------------------------------

# 2. Libraries and helpers

```{r libraries}
suppressPackageStartupMessages({
  library(tidycensus)
  library(tidyverse)
  library(sf)
  library(knitr)
  library(kableExtra)
  library(scales)
  library(showtext)
})

font_add_google("Exo 2",     "Exo 2")
font_add_google("Open Sans", "Open Sans")
showtext_auto()
showtext_opts(dpi = 150)

# Helper: safely convert columns that might be character (ACS suppression codes
# like "N", "-", "(X)", "**") to numeric.
safe_numeric <- function(x) suppressWarnings(as.numeric(as.character(x)))

# ---------------------------------------------------------------------------
# Table styling helpers
# ---------------------------------------------------------------------------

# Wrap a numeric score (0-100) in a color-coded HTML cell that matches the map
# gradient. Automatically picks white or dark text depending on background.
score_cell <- function(x, lo = min(x, na.rm = TRUE), hi = max(x, na.rm = TRUE)) {
  pal_fn <- scales::col_numeric(
    palette = c("#7aa88c", "#e7d9b8", "#d97757", "#a83c2b"),
    domain  = c(lo, hi)
  )
  vapply(x, function(v) {
    if (is.na(v)) return(as.character(NA))
    bg  <- pal_fn(v)
    txt <- if (v > 55) "white" else "#2b2b2b"
    wt  <- if (v >= 75) "bold" else "normal"
    kableExtra::cell_spec(
      round(v, 1),
      background = bg, color = txt, bold = (v >= 75),
      extra_css  = paste0("font-weight:", wt, "; padding: 4px 10px;")
    )
  }, character(1))
}

# Standard table wrapper: navy header, clean alternating rows, Exo 2 headings.
# Uses kableExtra::kbl() (not knitr::kable()) so that escape = FALSE reliably
# renders HTML from score_cell() as actual colored spans, not escaped text.
hvi_kable <- function(df, caption, escape = FALSE, full_width = FALSE, ...) {
  df |>
    kableExtra::kbl(caption = caption, escape = escape, align = "l",
                    format = "html", ...) |>
    kableExtra::kable_styling(
      bootstrap_options = c("hover", "condensed"),
      full_width  = full_width,
      font_size   = 13
    ) |>
    kableExtra::row_spec(
      0,
      bold       = TRUE,
      color      = "white",
      background = "#1f3b57",
      extra_css  = "font-family: 'Exo 2', sans-serif; font-size: 13px; letter-spacing: 0.03em;"
    )
}
```

------------------------------------------------------------------------

# 3. Pull ACS data

```{r pull-acs}
# Build a named vector of every ACS variable we need (numerators + denominators).
# Numerator field can hold a single ACS code OR a comma-separated list of codes
# to be summed (e.g. "B25014_005,B25014_006,B25014_011,B25014_012" for total
# overcrowded households across owner and renter cells).
acs_vars <- var_spec |>
  select(acs_numerator, acs_denominator) |>
  pivot_longer(everything(), values_to = "acs_code") |>
  filter(!is.na(acs_code)) |>
  pull(acs_code) |>
  strsplit(",") |>
  unlist() |>
  trimws() |>
  unique()

# Add LEP (B16004) cells for "speaks English less than very well" automatically.
if (use_lep) {
  b16004_vars <- load_variables(study_area$acs_year, study_area$acs_survey,
                                cache = FALSE) |>
    filter(str_detect(name, "^B16004_"))

  # --- Tier 1: look for aggregate "less than very well" rows.
  # The label format changed across ACS vintages; allow for curly/straight
  # quotes and extra whitespace between "less than" and "very well".
  lep_table <- b16004_vars |>
    filter(str_detect(
      label,
      regex("less.{0,5}than.{0,15}very.{0,5}well", ignore_case = TRUE)
    ))

  # --- Tier 2: if no aggregate rows found, use "not well" + "not at all" rows.
  # These are the two clearest LEP sub-categories. The "well" stratum is
  # technically LEP too but is harder to isolate without the aggregate row, so
  # we skip it in this fallback (slight undercount, acceptable for a small-area
  # index like this one).
  if (nrow(lep_table) == 0) {
    message("B16004: no 'less than very well' aggregate rows found — ",
            "falling back to 'not well' and 'not at all' detail rows (slight LEP undercount).")
    lep_table <- b16004_vars |>
      filter(str_detect(label, regex("\\bnot well\\b|not at all", ignore_case = TRUE)))
  }

  # --- Tier 3: if still empty, print diagnostics and skip LEP gracefully.
  if (nrow(lep_table) == 0) {
    warning("Could not identify LEP cells in B16004 for the ",
            study_area$acs_year, " ", study_area$acs_survey, " release. ",
            "Skipping LEP. First 10 B16004 labels:\n",
            paste(head(b16004_vars$label, 10), collapse = "\n"))
    lep_codes <- character(0)
  } else {
    lep_codes <- lep_table$name
    message("LEP: matched ", length(lep_codes), " B16004 cells.")
  }

  lep_universe <- "B16004_001"
  acs_vars     <- unique(c(acs_vars, lep_codes, lep_universe))
}

# Pull from the Census API. `geometry = TRUE` returns tract polygons.
raw_acs <- get_acs(
  geography = "tract",
  state     = study_area$state,
  county    = study_area$county,
  year      = study_area$acs_year,
  survey    = study_area$acs_survey,
  variables = acs_vars,
  geometry  = TRUE,
  cache_table = TRUE
)

# Filter to the study area tracts only.
seamaac <- raw_acs |>
  mutate(tract_code = str_sub(GEOID, -6)) |>
  filter(tract_code %in% study_area$tract_codes)
```

------------------------------------------------------------------------

# 4. Reshape long → wide

ACS data arrives in long format (one row per tract × variable). We spread it so each row is one tract and each column is one variable.

:::: callout
::: callout-title
Long vs. wide data
:::

"Long" means stacked — each row is one measurement. "Wide" means flat — each row is a whole neighborhood with all its measurements side-by-side. PCA needs the wide shape.
::::

```{r reshape-wide}
seamaac_wide <- seamaac |>
  st_drop_geometry() |>
  select(GEOID, NAME, tract_code, variable, estimate) |>
  pivot_wider(names_from = variable, values_from = estimate)

# Keep the tract geometry separately so we can rejoin it for mapping later.
tract_geom <- seamaac |>
  select(GEOID) |>
  distinct(GEOID, .keep_all = TRUE)
```

------------------------------------------------------------------------

# 5. Build derived variables (fully automatic)

This loop walks through `var_spec` and constructs every variable — no hardcoded formulas.

```{r build-derived}
# Start with GEOID + tract identifiers
derived <- seamaac_wide |>
  select(GEOID, NAME, tract_code) |>
  mutate(tract_short = str_remove(NAME, ";.*$"))

# Helper: pull a numeric column out of seamaac_wide by its ACS code.
get_col <- function(df, code) {
  if (is.na(code) || !(code %in% names(df))) return(rep(NA_real_, nrow(df)))
  safe_numeric(df[[code]])
}

# Helper: sum one or more ACS codes (comma-separated numerator field).
sum_cols <- function(df, codes_str) {
  if (is.na(codes_str)) return(rep(NA_real_, nrow(df)))
  codes <- trimws(strsplit(codes_str, ",")[[1]])
  mat   <- vapply(codes, function(c) get_col(df, c),
                  numeric(nrow(df)))
  if (is.null(dim(mat))) mat <- matrix(mat, ncol = 1)
  rowSums(mat, na.rm = TRUE)
}

# Build each variable from the spec.
for (i in seq_len(nrow(var_spec))) {
  row  <- var_spec[i, ]
  fn   <- row$friendly_name
  num  <- sum_cols(seamaac_wide, row$acs_numerator)
  den  <- if (!is.na(row$acs_denominator))
            sum_cols(seamaac_wide, row$acs_denominator) else NULL

  val <- switch(row$transform,
    "rate" = ifelse(!is.null(den) & !is.na(den) & den > 0, num / den, NA_real_),
    "log"  = ifelse(num > 0, log(num), NA_real_),
    "none" = num
  )
  derived[[fn]] <- val
}

# LEP rate — summed across all "less than very well" cells / total pop 5+
if (use_lep) {
  # Pull each LEP cell as its own column, then bind into a matrix.
  # vapply guarantees a numeric matrix even when there's only one code.
  lep_codes_present <- intersect(lep_codes, names(seamaac_wide))

  if (length(lep_codes_present) >= 1) {
    lep_mat <- vapply(lep_codes_present,
                      function(c) get_col(seamaac_wide, c),
                      numeric(nrow(seamaac_wide)))
    # vapply with a length-N output returns an N x K matrix when K >= 2,
    # but a length-N vector when K == 1. Force a matrix either way.
    if (is.null(dim(lep_mat))) lep_mat <- matrix(lep_mat, ncol = 1)

    lep_sum   <- rowSums(lep_mat, na.rm = TRUE)
    lep_total <- get_col(seamaac_wide, lep_universe)

    derived$pct_lep <- ifelse(!is.na(lep_total) & lep_total > 0,
                              lep_sum / lep_total, NA_real_)

    # Add LEP to the spec so it flows through the rest of the pipeline.
    var_spec <- bind_rows(var_spec, tibble(
      friendly_name   = "pct_lep",
      acs_numerator   = "B16004 (summed)",
      acs_denominator = "B16004_001",
      transform       = "rate",
      direction       = +1,
      domain          = "language",
      label           = "% population 5+ LEP"
    ))
  } else {
    message("No B16004 'less than very well' cells found in the data; skipping LEP.")
  }
}

derived
```

------------------------------------------------------------------------

# 6. Data validation (automatic)

We drop any variable that:

1.  Has **zero variance** (same value for every tract — can't contribute to PCA).
2.  Has **too many missing values** (more than half the tracts).

Anything dropped is reported so nothing is hidden.

```{r validate}
pca_candidates <- var_spec$friendly_name

var_check <- tibble(
  variable = pca_candidates,
  n_missing = sapply(pca_candidates,
                     function(v) sum(is.na(derived[[v]]))),
  variance  = sapply(pca_candidates,
                     function(v) {
                       x <- derived[[v]]
                       if (all(is.na(x))) return(NA_real_)
                       stats::var(x, na.rm = TRUE)
                     })
)

# Flag problems
var_check <- var_check |>
  mutate(
    status = case_when(
      is.na(variance)          ~ "dropped: all missing",
      variance == 0            ~ "dropped: zero variance",
      n_missing > nrow(derived)/2 ~ "dropped: too many missing",
      TRUE                     ~ "kept"
    )
  )

dropped_vars <- var_check |> filter(status != "kept") |> pull(variable)
kept_vars    <- var_check |> filter(status == "kept") |> pull(variable)

var_check |>
  mutate(
    variance  = round(variance, 4),
    status    = cell_spec(
      status,
      color = case_when(
        status == "kept"               ~ "#1f3b57",
        str_detect(status, "dropped") ~ "#a83c2b",
        TRUE                          ~ "#2b2b2b"
      ),
      bold = status == "kept"
    )
  ) |>
  hvi_kable("Table 1 — Variable health check", escape = FALSE)
```

:::: callout
::: callout-title
Why this matters
:::

PCA breaks when a variable has the same value for every neighborhood (there's nothing to compare) or when most of the data is missing. Dropping those variables automatically keeps the math clean — and it's transparent, because Table 1 shows exactly what was removed and why.
::::

```{r filter-kept}
# Subset to kept variables
active_spec <- var_spec |> filter(friendly_name %in% kept_vars)

# For PCA we need complete cases. Impute remaining NAs with the column median.
pca_input <- derived |>
  select(GEOID, NAME, tract_short, tract_code, all_of(kept_vars)) |>
  mutate(across(all_of(kept_vars),
                ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)))
```

------------------------------------------------------------------------

# 7. Two separate PCAs

The index is built in two stages, because housing/economic stress and language access measure different things.

:::: callout
::: callout-title
What PCA does
:::

Principal Component Analysis finds the direction in the data where neighborhoods differ from each other the most. The first principal component (PC1) is that biggest-difference axis. PC2 is the next biggest, and so on. Each PC is a weighted mix of the original variables.
::::

```{r run-pcas}
# ---- Helper: run a PCA on a subset of variables, orient signs automatically.
# Gracefully handles single-variable domains by returning a direction-flipped
# z-score in the shape of a PCA result, so the rest of the pipeline works the
# same way whether a domain has 1 variable or 15.
run_oriented_pca <- function(pca_input, spec_subset) {
  vars <- spec_subset$friendly_name
  if (length(vars) == 0) return(NULL)

  X        <- pca_input |> select(all_of(vars)) |> as.matrix()
  X_scaled <- scale(X)
  direction <- spec_subset$direction[match(vars, spec_subset$friendly_name)]

  # ---- Single-variable domain: return direction-flipped z-score as the score.
  if (length(vars) == 1) {
    s <- as.numeric(X_scaled) * direction
    rot <- matrix(direction, nrow = 1, ncol = 1,
                  dimnames = list(vars, "PC1"))
    return(list(
      pca      = list(rotation = rot, x = matrix(s, ncol = 1,
                                                 dimnames = list(NULL, "PC1"))),
      scores   = data.frame(PC1 = s),
      variance = c(PC1 = 1.0),
      proxy    = s
    ))
  }

  # ---- Multi-variable domain: run PCA and auto-orient each PC.
  pca <- prcomp(X_scaled, center = FALSE, scale. = FALSE)

  # Theoretical vulnerability proxy: sum of z-scores weighted by each
  # variable's direction (+1 = higher is worse; -1 = higher is better). This setup forces the direction into an explicit kx1 matrix regardless of how it came out of the tibble subset. 
  proxy <- as.numeric(X_scaled %*% matrix(direction, ncol = 1))

  # For every PC, flip its sign so that it correlates positively with the
  # proxy. That way higher PC score = more vulnerable, automatically.
  scores <- pca$x
  for (k in seq_len(ncol(scores))) {
    r <- suppressWarnings(stats::cor(scores[, k], proxy))
    if (!is.na(r) && r < 0) {
      scores[, k]        <- -scores[, k]
      pca$rotation[, k]  <- -pca$rotation[, k]
    }
  }

  list(
    pca      = pca,
    scores   = as.data.frame(scores),
    variance = summary(pca)$importance["Proportion of Variance", ],
    proxy    = proxy
  )
}

# Housing & Economic PCA
he_spec <- active_spec |> filter(domain == "housing_econ")
he_out  <- run_oriented_pca(pca_input, he_spec)

# Language PCA
lang_spec <- active_spec |> filter(domain == "language")
lang_out  <- run_oriented_pca(pca_input, lang_spec)
```

:::: callout
::: callout-title
How we decide which direction is "more vulnerable"
:::

We don't eyeball the PCA output. Instead, every variable in the spec carries a `direction` flag (+1 if higher = more vulnerable, −1 if higher = less vulnerable). The code builds a theory-based "vulnerability proxy" from those flags, and for each PC it checks whether the PC's scores move *with* the proxy or against it. If the PC moves against the proxy, we flip its sign. This runs automatically — there is no place in the document where anyone has to set a sign by hand.
::::

------------------------------------------------------------------------

# 8. Sub-index and final HVI construction

```{r build-composite}
# Weighted PC score for a sub-index. Uses enough PCs to cover `var_threshold`
# of the total variance, or all of them if there are fewer.
weighted_pc_score <- function(out, var_threshold = 0.80) {
  if (is.null(out)) return(NULL)
  cumv <- cumsum(out$variance)
  k    <- max(1, which(cumv >= var_threshold)[1])
  if (is.na(k)) k <- length(out$variance)
  wts  <- out$variance[1:k]
  wts  <- wts / sum(wts)
  s    <- as.matrix(out$scores[, 1:k, drop = FALSE]) %*% wts
  as.numeric(s)
}


# Construct raw and 0–100 sub-index scores
he_raw   <- weighted_pc_score(he_out)
lang_raw <- weighted_pc_score(lang_out)

# Normalize subindex_weights so they sum to 1
w <- subindex_weights / sum(subindex_weights)

scores <- pca_input |>
  select(GEOID, NAME, tract_short, tract_code) |>
  mutate(
    housing_econ_score = if (!is.null(he_raw))   he_raw   else NA_real_,
language_score     = if (!is.null(lang_raw)) lang_raw else NA_real_)|>
  mutate(
    # Final HVI is the weighted average of the two 0-100 sub-indices.
    # We deliberately do NOT rescale the final score to 0-100 again — that
    # rescale forces the highest tract to 100 even if it isn't extreme on
    # both sub-indices, which makes the final map look identical to whichever
    # sub-index dominates. Leaving it as a true weighted average preserves
    # the contribution of each sub-index.
    hvi_score = case_when(
      !is.na(housing_econ_score) & !is.na(language_score) ~
        housing_econ_score * w["housing_econ"] + language_score * w["language"],
      !is.na(housing_econ_score) ~ housing_econ_score,
      !is.na(language_score)     ~ language_score,
      TRUE ~ NA_real_
    )
  ) |>
  # Determine which sub-index dominates each tract for the analysis section.
  mutate(
    dominant_subindex = case_when(
      is.na(housing_econ_score) | is.na(language_score) ~ NA_character_,
      housing_econ_score - language_score >  10 ~ "housing/economic",
      language_score - housing_econ_score >  10 ~ "language access",
      TRUE                                       ~ "balanced"
    ),
    hvi_rank = rank(-hvi_score, ties.method = "min")
  ) |>
  arrange(desc(hvi_score))
```

------------------------------------------------------------------------

# 9. Results

## Table 2 — Variance explained by each PCA

```{r tbl-variance}
bind_rows(
  if (!is.null(he_out))   tibble(component = names(he_out$variance),
                                 domain    = "Housing & Economic",
                                 prop_var  = he_out$variance) else NULL,
  if (!is.null(lang_out)) tibble(component = names(lang_out$variance),
                                 domain    = "Language Access",
                                 prop_var  = lang_out$variance) else NULL
) |>
  mutate(`Variance explained` = percent(prop_var, accuracy = 0.1)) |>
  select(Domain = domain, Component = component, `Variance explained`) |>
  hvi_kable("Table 2 — How much each principal component explains")
```

## Table 3 — Variable loadings on PC1 of each PCA

```{r tbl-loadings}
make_loading_tbl <- function(out, spec_subset, domain_label) {
  if (is.null(out)) return(NULL)
  rot <- out$pca$rotation
  ncp <- min(3, ncol(rot))
  as.data.frame(rot[, 1:ncp, drop = FALSE]) |>
    tibble::rownames_to_column("friendly_name") |>
    left_join(spec_subset |> select(friendly_name, label, direction),
              by = "friendly_name") |>
    mutate(
      Direction = ifelse(direction == 1, "higher = more vulnerable",
                                         "higher = less vulnerable"),
      Domain    = domain_label
    ) |>
    select(Domain, Variable = label, Direction, everything(),
           -friendly_name, -direction) |>
    arrange(desc(abs(PC1)))
}

loadings_he   <- make_loading_tbl(he_out,   he_spec,   "Housing & Economic")
loadings_lang <- make_loading_tbl(lang_out, lang_spec, "Language Access")

bind_rows(loadings_he, loadings_lang) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  hvi_kable("Table 3 — How each variable contributes to the top PCs")
```

## Table 4 — Final HVI scores

```{r tbl-hvi}
scores |>
  transmute(
    Rank  = hvi_rank,
    Tract = tract_short,
    `Housing & Econ`  = score_cell(housing_econ_score),
    `Language Access` = score_cell(language_score),
    `Final HVI`       = score_cell(hvi_score),
    `Dominant driver` = dominant_subindex
  ) |>
  hvi_kable(
    "Table 4 — Sub-index and combined HVI scores (higher = more vulnerable)",
    escape = FALSE
  ) |>
  kableExtra::column_spec(5, bold = TRUE, border_left = TRUE) |>
  kableExtra::column_spec(1, width = "3em", color = "#4a7ca0", bold = TRUE)
```

## Table 5 — Indicator snapshot

```{r tbl-indicators}
pca_input |>
  select(Tract = tract_short, all_of(kept_vars)) |>
  left_join(scores |> select(tract_short, hvi_rank, hvi_score),
            by = c("Tract" = "tract_short")) |>
  arrange(hvi_rank) |>
  select(Rank = hvi_rank, Tract, -hvi_score, everything()) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  hvi_kable("Table 5 — All indicators by tract (sorted by final HVI, highest vulnerability first)",
            full_width = TRUE) |>
  kableExtra::column_spec(1, color = "#4a7ca0", bold = TRUE) |>
  kableExtra::scroll_box(width = "100%", height = "420px")
```

------------------------------------------------------------------------

# 10. Maps

```{r maps-join}
mapped <- tract_geom |>
  left_join(scores |> select(GEOID, housing_econ_score,
                             language_score, hvi_score, tract_short),
            by = "GEOID")
```

```{r map-helper}
draw_choropleth <- function(sf_df, fill_col, title, subtitle,
                            caption, label_col = "tract_short") {
  ggplot(sf_df) +
    geom_sf(aes(fill = .data[[fill_col]]),
            color = "white", linewidth = 0.6) +
    scale_fill_gradientn(
      colors  = hvi_gradient,
      name    = "Score",
      labels  = label_number(accuracy = 1),
      guide   = guide_colorbar(barwidth = 10, barheight = 0.7,
                               title.position = "top")
    ) +
    ggrepel::geom_label_repel(
  aes(geometry = geometry,
      label = paste0(.data[[label_col]], "\n", round(.data[[fill_col]], 1))),
  stat = "sf_coordinates",
  size = 2.8, fill = "white", alpha = 0.85,
  family = "Open Sans", color = pal$navy,
  label.size = 0.2,
  min.segment.length = 0,
  max.overlaps = Inf
) +
    labs(title = title, subtitle = subtitle, caption = caption) +
    theme_void(base_family = "Open Sans") +
    theme(
      plot.title    = element_text(family = "Exo 2", face = "bold",
                                   size = 16, color = pal$navy),
      plot.subtitle = element_text(family = "Open Sans",
                                   size = 10, color = "grey35"),
      plot.caption  = element_text(family = "Open Sans",
                                   size = 8, color = "grey45",
                                   hjust = 0),
      legend.position   = "bottom",
      legend.title      = element_text(family = "Exo 2",
                                       size = 9, color = pal$navy),
      legend.text       = element_text(size = 8),
      plot.margin       = margin(12, 12, 12, 12)
    )
}
```

## Map 1 — Final Heat Vulnerability Index

```{r map-final}
draw_choropleth(
  mapped, "hvi_score",
  title    = "SEAMAAC Heat Vulnerability Index",
  subtitle = paste0(study_area$acs_year - 4, "–", study_area$acs_year,
                    " ACS 5-Year · ", nrow(scores), " census tracts"),
  caption  = "Higher scores indicate greater heat vulnerability.\nCombines a Housing & Economic sub-index with a Language Access sub-index."
)
```

## Map 2 — Housing & Economic sub-index

```{r map-housing}
draw_choropleth(
  mapped, "housing_econ_score",
  title    = "Housing & Economic Sub-Index",
  subtitle = "Drivers: cost burden, crowding, housing quality, poverty, income",
  caption  = "Higher scores indicate greater housing and economic stress."
)
```

## Map 3 — Language Access sub-index

```{r map-language}
draw_choropleth(
  mapped, "language_score",
  title    = "Language Access Sub-Index",
  subtitle = "Drivers: Limited English Proficiency and linguistic isolation",
  caption  = "Higher scores indicate greater language-access barriers during heat events."
)
```

------------------------------------------------------------------------

# 11. Results and interpretation

```{r analysis-helpers, include = FALSE}
# Utilities that make the narrative paragraphs below programmatic.

# Which two variables most strongly pull a tract up relative to the others?
# We compute a z-score for each indicator, flip by direction, and return the
# indicators where this tract is most above average.
top_drivers_for_tract <- function(tract_code_val, n = 2) {
  tract_row <- pca_input |> filter(tract_code == tract_code_val)
  if (nrow(tract_row) == 0) return(character(0))
  z <- sapply(active_spec$friendly_name, function(v) {
    x  <- pca_input[[v]]
    mu <- mean(x, na.rm = TRUE)
    sd <- stats::sd(x, na.rm = TRUE)
    if (is.na(sd) || sd == 0) return(NA_real_)
    direction <- active_spec$direction[active_spec$friendly_name == v]
    ((tract_row[[v]] - mu) / sd) * direction
  })
  z <- sort(z, decreasing = TRUE, na.last = NA)
  idx   <- head(names(z), n)
  active_spec$label[match(idx, active_spec$friendly_name)]
}

# Pearson correlation between housing and language scores (how concordant are
# the two sub-indices?).
he_la_corr <- suppressWarnings(
  cor(scores$housing_econ_score, scores$language_score,
      use = "complete.obs")
)

scores_for_text <- scores |>
  arrange(desc(hvi_score)) |>
  mutate(rank = row_number())
```

## 11.1 Which tracts are most vulnerable?

```{r top-bottom-narrative, results = "asis"}
top_tract    <- scores_for_text |> slice(1)
bottom_tract <- scores_for_text |> slice(n())

top_drivers    <- top_drivers_for_tract(top_tract$tract_code, 2)
bottom_drivers <- top_drivers_for_tract(bottom_tract$tract_code, 2)

cat(sprintf(
"The census tract scoring highest on the combined HVI is **%s** (HVI = %.1f, with 0 being the rough average). The two indicators pulling this tract up the most are *%s* and *%s*. ",
  top_tract$tract_short, top_tract$hvi_score,
  top_drivers[1], top_drivers[2]
))

cat(sprintf(
"The tract scoring lowest is **%s** (HVI = %.1f). Its lower score reflects relatively better conditions on *%s* and *%s* compared with the rest of the study area. ",
  bottom_tract$tract_short, bottom_tract$hvi_score,
  bottom_drivers[1], bottom_drivers[2]
))

cat(sprintf(
"The full ranking (highest to lowest) is: **%s**.",
  paste(scores_for_text$tract_short, collapse = " > ")
))
```

## 11.2 Do housing stress and language access point to the same tracts?

```{r concordance-narrative, results = "asis"}
corr_text <- case_when(
  is.na(he_la_corr)         ~ "undefined (one sub-index is constant)",
  abs(he_la_corr) < 0.3     ~ sprintf("weak (r = %.2f)",     he_la_corr),
  abs(he_la_corr) < 0.7     ~ sprintf("moderate (r = %.2f)", he_la_corr),
  TRUE                      ~ sprintf("strong (r = %.2f)",   he_la_corr)
)

interpretation <- if (is.na(he_la_corr)) {
  "we cannot compare the two sub-indices directly."
} else if (he_la_corr > 0.7) {
  "the same tracts that are most housing-stressed also have the greatest language-access barriers — so the final map closely resembles either sub-index map on its own."
} else if (he_la_corr > 0.3) {
  "the two sub-indices agree on most tracts but diverge on some. The divergent tracts are where the combined index adds the most value relative to housing-only measures."
} else if (he_la_corr > -0.3) {
  "the two sub-indices capture largely different signals across tracts. The combined HVI is a genuine blend of both, and divergent tracts stand out clearly."
} else {
  "the two sub-indices trend in opposite directions — some of the housing-stressed tracts are actually better off on language access, and vice versa. This is a strong case for reporting them separately."
}

cat(sprintf(
"The correlation between the Housing & Economic and Language Access sub-indices is **%s**, meaning %s",
  corr_text, interpretation
))
```

```{r concordance-table}
scores |>
  transmute(
    Rank             = hvi_rank,
    Tract            = tract_short,
    `Housing & Econ` = score_cell(housing_econ_score),
    `Language Access`= score_cell(language_score),
    `Difference (H − L)` = round(housing_econ_score - language_score, 1),
    `Dominant driver`= dominant_subindex,
    `Final HVI`      = score_cell(hvi_score)
  ) |>
  arrange(Rank) |>
  hvi_kable(
    "Table 6 — Sub-index comparison. A large positive difference means housing stress is dominant; negative means language barriers are dominant.",
    escape = FALSE
  ) |>
  kableExtra::column_spec(1, color = "#4a7ca0", bold = TRUE) |>
  kableExtra::column_spec(7, bold = TRUE, border_left = TRUE)
```

## 11.3 Tract-by-tract summary

```{r per-tract-summary, results = "asis"}
for (i in seq_len(nrow(scores_for_text))) {
  t  <- scores_for_text[i, ]
  td <- top_drivers_for_tract(t$tract_code, 2)
  dom_text <- case_when(
    t$dominant_subindex == "housing/economic" ~
      "Housing and economic stress is the primary driver here.",
    t$dominant_subindex == "language access"  ~
      "Language-access barriers are the primary driver here.",
    t$dominant_subindex == "balanced"         ~
      "Housing and language dimensions contribute roughly equally.",
    TRUE ~ ""
  )
  cat(sprintf(
"- **%s (Rank %d, HVI %.1f)** — H&E %.1f · Lang %.1f. %s Strongest indicators: *%s*; *%s*.\n\n",
    t$tract_short, t$rank, t$hvi_score,
    t$housing_econ_score, t$language_score,
    dom_text, td[1], td[2]
  ))
}
```

:::: callout
::: callout-title
How to read this
:::

"Rank" orders tracts from most to least heat-vulnerable in this study area. "H&E" is the housing & economic sub-index and "Lang" is the language access sub-index. The "strongest indicators" for each tract are the variables where that tract sits farthest above the local average (after flipping direction so that higher always means more vulnerable). They explain *why* the tract scored where it did.
::::

------------------------------------------------------------------------

# 12. What this index adds beyond Philadelphia's existing HVI

The Philadelphia city-wide HVI focuses primarily on demographic and surface-temperature factors. The index in this report is designed to complement it with indicators especially relevant to SEAMAAC's community:

-   **Language access.** LEP population share and linguistic isolation at the household level — critical for outreach, cooling-center awareness, and emergency communications.
-   **Household-level rent burden.** Share of renters paying **50% or more** of income on rent (severe cost burden), which strains the ability to afford utilities including air conditioning.
-   **Occupants per room.** Crowding measured properly (not just average household size) — a stronger signal of heat stress in housing.
-   **Heating / housing infrastructure gaps.** Share of units with no heating fuel and mobile-home / non-standard structures.
-   **Housing instability.** Share of recent movers, who are more likely to be in unstable situations and less likely to know local cooling resources.

------------------------------------------------------------------------

# 13. Limitations

-   **Small n.** With only `r nrow(scores)` tracts, PCA is sensitive; use these scores as a relative ranking for planning, not as precise estimates.
-   **AC not measured at the tract level.** The ACS does not ask about air conditioning at tract resolution, so direct cooling-equipment data is unavailable.
-   **ACS 5-year estimates are averaged** over five years; they can mask rapid neighborhood change.
-   **Imputation.** Any missing tract values are replaced with the column median. Table 1 reports how many values were imputed per variable.

------------------------------------------------------------------------

# 14. Reproducibility footer

```{r session-info}
sessionInfo()
```
