---
title: "Measuring Occupational Heat Exposure at the Local Level"
author: "Sydney Jones, BSW"
bibliography: references.bib
format:
  html:
    toc: true
    toc-location: right        # floating TOC
    toc-depth: 3
    number-sections: true
    code-summary: "Click to view code"
    theme: litera               # clean default
    anchor-sections: true # optional custom tweaks
    df-print: paged
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
  

---

# Overview

Adapted from methodology outlined by Zoey Laskaris in their article "Occupational characteristics are missing from heat vulnerability indices: a study in New York and new jersey" (2005).

# Setting Up R

## Options

```{r setup}
options(
  tigris_use_cache = T,
  scipen = 999
)
```

## Packages

```{r packages}

library(pacman)

p_load(tidyverse, sf, tigris, tidycensus, data.table, DT, tmap, stringr, kableExtra)
```

# Pulling Data from ACS

## Create Filtering Variables

To filter by census tract:

```{r tracts}

tracts <- c("004001", "004002", "004101", "004103", "004104", "004201", "004202", "037200")

```

## Pulling Economic Characteristics (DP03) From ACS

### Pulling Variable Labels

When you initially pull the DPO3 table, the data will not include human-readable lables. We pull the variable labels separately so that we can have labels and aren't working with basic codes.

We start with the labels for all profile table variables.

```{r vars}

vars <- load_variables(2024, "acs5/profile", cache = TRUE)

```

Filter for the table we are looking for, `DP03 Selected Economic Characteristics`.

```{r dp03_vars}

dp03_vars <- vars %>%
  filter(grepl("^DP03", name)) %>%
  select(name, label)
```

## Pulling Economic Characteristics Table

Next, use tidycensus's `get_acs()` call to pull county-level information from `DP03`. [^1]

In the included chunk, I pulled the stored data from my local memory, but the API call is commented out and ready to be used. 

[^1]: I've used text strings to indicate state and county, but you can also use the FIPS code, a 5-digit identifier system to signify state and county developed by the US Census Bureau. FIPS codes are made of a 2-digit number representing the state and a 3-digit number representing a county within that state, arranged in alphabetical order. (ex: PA = 42, county = 101). FIPS codes are combined with identifiers of smaller units of geography to create the GEOID of a specific place.

    A list of state- and county-level FIPS codes can be found here: <https://transition.fcc.gov/oet/info/maps/census/fips/fips.txt>.

    More information on geographic identifiers, including FIPS codes and others, can be found here: <https://www.census.gov/programs-surveys/geography/guidance/geo-identifiers.html>.

```{r dp03}

dp03 <- readRDS(here::here("occupational_exposure/data/raw/dp03.rds"))

# dp03 <- get_acs(
#   geography = "tract", #specifies the geographic unit of analysis
#   state = "PA", #state of interest
#   county = "Philadelphia County", #county of interest
#   year = 2024, #survey year
#   survey = "acs5", #specific survey, ACS 5-Year
#   table = "DP03", #table we are pulling
#   cache_table = T, #this caches the results so we can pull it faster next time
#   geometry = T, #tells tidycensus that we want to include polygon shapes as well for mapping later
#   progress_bar = F #this is crucial!! Keeps all the extra progress junk from printing out in report later
# ) 


```

To make it easier to bring in our variable labels, we rename the `label` column they live under in our `dp03_vars` table to `variable` so it matches that present in our `dp03` table.

```{r rename_vars}

dp03_vars <- dp03_vars %>% 
  rename(variable = name)

```

## Join Variable Labels to Data

Perform a `left_join()` so that we have human-readable labels. A left join combines our first (left) df, `dp03`, with a new df `dp03_vars` based on values present in a column that we specify with the argument `by =` . When choosing the joining value, we want to make sure it is a unique charachteristic of the data it represents. This basically means that we want to pick something that is only present one time for each item (row) in the data.

It is called a *left* join because it keeps everything from the df on the left, which in this instance we want to be our df with county-level data. Anything present in `dp03_vars` that doesn't match is left behind. There are several other join types. Left is the most common.

```{r dp03_join}

dp03_join <- dp03 %>% 
  left_join(dp03_vars, by = "variable")

```

# Cleaning Data

## Coding for Tract and Subsetting our Study Area

The data we pulled doesn't have a dedicated column for the census tract code. That is ok–it lives inside of the GEOID. Subset the census tract code, which is the last six digits of the GEOID (the first five being the FIPS representing county and state) into its own column.

```{r tract_column}

dp03_join <- dp03_join %>% 
  mutate(tract = str_sub(GEOID, -6))

```

Now we subset the data and remove only the information on the census tracts that are within our desired study area. We use the `tracts` vector we created earlier that contains the six-digit census tract codes.

```{r study_dp03}

study_dp03 <- dp03_join %>% 
  filter(tract %in% tracts)
```

The data includes estimates, margins of error, and percentages. we will take only the estimates, since we will be calculating our own percentage shares later. We will also break up the column that names the geography.

```{r study_dp03e}

study_dp03e <- study_dp03 %>% 
  filter(!str_detect(variable, "P$")) %>%  #estimate percentages are denoted by a P at the end of the variable code
  separate(
    col = NAME,
    into = c("tract_name", "county_name", "state_name"),
    sep = "; ",
    remove = T #we don't need to hold on to this since it lives in the previous df
  )
```

Next, we break up the weird and long `label` column into separate columns so that it is easier to read and understand what exactly each row's estimate represents. Similar to the previous step, we do not need to hold on to the original column since it lives in `study_dp03`.

```{r separate_labels}
study_dp03e <- study_dp03e %>% 
  separate(
    col = label,
    into = c("measure", "indicator", "metric", "detail_1", "detail_2", "detail_3", "detail_4"),
    sep = "!!",
    remove = T
      )
```

### Examination of Different Categories

Run a few summary tables to see what the different categories within the data represent.

Summarizing the `indicator` column shows us what different categories of information are included within `DP03`.

```{r indicator_sum}

indicator_sum <- study_dp03e %>% 
  group_by(indicator) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  arrange(desc(n))

indicator_sum %>%
  rename("Indicator Category" = indicator,
         "Number of Observations" = n) %>% 
  st_drop_geometry() %>% 
  kbl(caption = "Count of Observations by Indicator") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )  


```

A summary by `metric` shows us the count of each type of measured group across the data. For example, different units of measure included are workers, population members, non-family households, and even median earning values.

```{r metric_sum}

 metric_sum <- study_dp03e %>% 
   group_by(metric) %>% 
   summarise(n = n(), .groups = "drop") %>% 
  arrange(desc(n))
 
 metric_sum %>% 
   rename("Observation Metric" = metric,
          "Count of Observations by Metric" = n) %>% 
   st_drop_geometry() %>% 
  kbl(caption = "Count of Observations by Metric") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )  
 

```

We can summarize by both `indicator` and `metric` to see which indicators use different units of measure.

For example, when looking at income and benefits, there are 91 metrics reported related to families.

```{r indicator_metric_sum}

 indicator_metric_sum <- study_dp03e %>% 
   group_by(indicator, metric) %>% 
   summarise(n = n(), .groups = "drop") %>% 
  arrange(indicator, desc(n))
 
 
 indicator_metric_sum %>%
   rename("Measure Indicator" = metric,
          "Count of Observations by Indicator" = n) %>% 
   st_drop_geometry() %>% 
  kbl(caption = "Count of Observations by Metric") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )  
 


```

## Subsetting by Industry/Occuation

Subsetting a new df that includes sole information on industry and occupation. We also leave behind some unnecessary columns that we don't need right now. This provides us with a df of estimates for each industry and occupation by study area census tract.

Many of the following operations can be modified and used to analyze any of the other indicators available by modifying the indicator used in the filter filter call.

```{r ind_occ}

ind_occ <- study_dp03e %>% 
  filter(indicator == "INDUSTRY" | indicator == "OCCUPATION") %>% 
   select(
     GEOID,
     tract,
     tract_name,
     indicator,
     metric,
     industry_occupation = detail_1,
     estimate,
     moe,
     geometry
   ) %>% #we fill in the empty cells here with a new label because they represent total laborforce within the study area
   mutate(industry_occupation = case_when(
     is.na(industry_occupation) ~ "TOTAL STUDY AREA LABORFORCE",
     TRUE ~ industry_occupation
   )) 
```

## Summarize aggregate industry and occupation estimates 

Summarize the industry and occupation estimates for each census tract.

### Tract Level

We will clean up the `industry_occupation` column so that everything is identical – capitalized and without punctuation. We will also need to combine a couple of categories to match the computation that was done to calculate the indices.

```{r ind_occ_tract}

ind_occ_tract <- ind_occ %>% 
    mutate(
    industry_occupation = toupper(industry_occupation)
  ) %>% 
  mutate(
    industry_occupation = str_remove_all(industry_occupation, "[[:punct:]]")
  ) %>% 
  mutate( #combine retail and arts to accommodate the measure I created
    industry_occupation = case_when(
      industry_occupation == "RETAIL TRADE" ~ 
        "ARTS ENTERTAINMENT AND RECREATION AND ACCOMMODATION AND FOOD SERVICES AND RETAIL TRADE",
      industry_occupation == "ARTS ENTERTAINMENT AND RECREATION AND ACCOMMODATION AND FOOD SERVICES" ~ "ARTS ENTERTAINMENT AND RECREATION AND ACCOMMODATION AND FOOD SERVICES AND RETAIL TRADE",
      TRUE ~ industry_occupation
    )
  ) %>% 
  filter(
    industry_occupation != "TOTAL STUDY AREA LABORFORCE"
    )
```

Create estimate counts of employment by industry and occupation for each tract, and then employment for the total tract.

```{r ind_occ_tract_sum}

ind_occ_tract <- ind_occ_tract %>% 
   group_by(tract_name, indicator, industry_occupation) %>% 
   summarise(ind_occ_employment = sum(estimate)) %>% 
  ungroup() %>% 
  group_by(tract_name, indicator) %>% 
     mutate(
       total_tract_employment = sum(ind_occ_employment),
       .before = geometry
       ) %>% 
  ungroup()


```

Create a column for total study area employment.

```{r total_studyarea_employment}

ind_occ_tract <- ind_occ_tract %>% 
  group_by(indicator) %>% 
  mutate(
    total_studyarea_employment = sum(ind_occ_employment),
    .before = geometry) %>% 
  ungroup()
```

Make a table
```{r tract_level_summary}

ind_occ_tract %>% 
   mutate(
    tract_share = ind_occ_employment / total_tract_employment
  ) %>%
    arrange(tract_name, indicator, industry_occupation, ind_occ_employment, total_tract_employment, total_studyarea_employment, tract_share) %>% 
  rename(
    "Census Tract" = tract_name,
    "Industry or Occupation" = indicator,
    "Category" = industry_occupation,
    "Total Employees" = ind_occ_employment,
    "Total Employees by Tract" = total_tract_employment,
    "Total Study Area Employees" = total_studyarea_employment,
    "Share of Tract Employment" = tract_share
  ) %>%
  mutate(`Share of Tract Employment` = scales::percent(`Share of Tract Employment`, accuracy = 0.1)) %>%
  st_drop_geometry() %>% 
  kbl(caption = "Employment by Industry, Occupation, and Census Tract") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) 

```

### Study Area Level

We then combine the counts for all tracts within the entire study area so we can also analyze the neighborhood as a whole. During this step, we will also compute the share each industry and occupation occupies within the study area.

```{r ind_occ_agg}

ind_occ_agg <- ind_occ %>% 
    mutate(
    industry_occupation = toupper(industry_occupation)
  ) %>% 
  mutate(
    industry_occupation = str_remove_all(industry_occupation, "[[:punct:]]")
  ) %>% 
  mutate( #combine retail and arts to accommodate the measure I created
    industry_occupation = case_when(
      industry_occupation == "RETAIL TRADE" ~ 
        "ARTS ENTERTAINMENT AND RECREATION AND ACCOMMODATION AND FOOD SERVICES AND RETAIL TRADE",
      industry_occupation == "ARTS ENTERTAINMENT AND RECREATION AND ACCOMMODATION AND FOOD SERVICES" ~ "ARTS ENTERTAINMENT AND RECREATION AND ACCOMMODATION AND FOOD SERVICES AND RETAIL TRADE",
      TRUE ~ industry_occupation
    )
  ) %>% 
  filter(industry_occupation != "TOTAL STUDY AREA LABORFORCE")



```

Create summaries grouped by industry and occupation.

```{r ind_occ_agg_sum}

ind_occ_agg <- ind_occ_agg %>% 
   group_by(indicator, industry_occupation) %>% 
   summarise(ind_occ_employment = sum(estimate)) %>% 
  ungroup() %>% 
  group_by(indicator) %>% 
     mutate(
       total_studyarea_employment = sum(ind_occ_employment),
       .before = geometry
       ) %>% 
  ungroup()

```

We now have two summary panels: `ind_occ_tract`, which shows us the number of employees that work in each industry and occupation category; and `ind_occ_agg` that displays employment by each sector and industry across our entire study area.

## Summarize Tract and Study Area Laborforce

Next we create a table of the study area's laborforce, or its entire population of employable persons over the age of 16. This shows us the total number of eligible workers in each census tract.

We will rename the `estimate` and `moe` columns so that if we join this to our other data, we will know which is which.

```{r laborforce}

 laborforce <- study_dp03e %>% 
   filter(metric == "Civilian labor force") %>% 
   select(
     GEOID,
     tract_name,
     variable,
     indicator,
     metric,
     detail_1,
     estimate,
     moe
     ) %>% #filter out anything that has an NA estimate
   filter(
     !is.na(estimate)
   ) %>% #add a column that includes the total labor force for the entire study area
   group_by(tract_name, indicator, industry_occupation = metric) %>% 
   summarise(
     tract_laborforce = sum(estimate)
   ) %>% 
   ungroup() %>% 
   group_by(indicator) %>% 
   mutate(total_studyarea_laborforce = sum(tract_laborforce),
          .before = geometry) %>% 
   ungroup() %>%
   rename(sector = industry_occupation) %>% 
   st_drop_geometry()
```

## Join employment and laborforce data

```{r emp_labor}

emp_labor_tract <- ind_occ_tract %>% 
  left_join(laborforce, by = "tract_name") %>% 
  # reorganize columns for easier reading
  relocate(
    tract_laborforce, .after = total_tract_employment
    ) %>% 
  relocate(
    total_studyarea_laborforce, .after = total_studyarea_employment
  ) %>% 
  relocate(
    geometry, .after = last_col())

```

## Calculate unemployment rate and employment shares

### By tract

```{r tract_unemployment}

emp_labor_tract <- emp_labor_tract %>% 
  group_by(tract_name) %>% 
  mutate(
    tract_unemployment = 1 - (total_tract_employment / tract_laborforce)) %>% 
  ungroup() %>% 
  mutate(
    studyarea_unemployment = 1 - (total_studyarea_employment / total_studyarea_laborforce),
    tract_ind_occ_share = ind_occ_employment / total_tract_employment,
    .before = geometry
  )
```

### By Study Area

```{r labor_force_summary}
#create single vector so you can mutate it on to the ind_occ_agg
laborforce_summary <- laborforce %>% 
  select(indicator, total_studyarea_laborforce) %>% 
  distinct()

emp_labor <- ind_occ_agg %>% 
  mutate(
    
  ) %>% 
  mutate(
    ind_occ_employment_share = ind_occ_employment / total_studyarea_employment,
    total_studyarea_laborforce = laborforce_summary$total_studyarea_laborforce,
         studyarea_unemployment = 1 - (total_studyarea_employment / total_studyarea_laborforce),
         .before = geometry)
  
```

# Importing Industry/Occupation Heat Exposure Table

## Importing/Cleaning Data

Importing in the aggregated heat exposures table with means calculated based on the ACS categories.

```{r heat}

heat <- readxl::read_xlsx(here::here(("occupational_exposure/data/silver/acs_grouping_exposure.xlsx")))

```

Clean data so that it can easily be matched with other dfs, remove any duplicates and categories with NAs. ONET only has exposure measures for private industry jobs, so we want to take out anything that it did not provide measures for.

```{r heat_clean}

heat_clean <- heat %>% 
  mutate(
    industry_occupation = toupper(industry_occupation)
  ) %>% 
  mutate(
    industry_occupation = str_remove_all(industry_occupation, "[[:punct:]]")
  ) %>% 
  filter(
  !industry_occupation %in% c("PUBLIC ADMINISTRATION", "TOTAL STUDY AREA LABORFORCE")) %>% 
  distinct()

```

# Creating Indices

## Calculate Average, Weighted, and Standardized Scores in `heat_clean` df

Calculating different standardized exposure scores for each industry and occupation. 

These measures will be assigned to each industry or occupation moving forward. 

Also, removing `very_hot_very_cold` from the calculation bc it also includes extremely cold temperatures

```{r calculate_indices}

heat_clean <- heat_clean %>% 
  mutate( #weighted exposure
    weighted_exposure = (
      0.3 * indoors_controlled +
        0.4 * outdoors_exposed +
        0.3 * outdoors_undercover)
  ) %>% 
  mutate( #average exposure
    total_exposure = rowMeans(
      select(., indoors_controlled, outdoors_exposed, outdoors_undercover),
      na.rm = TRUE
    )
  )


heat_clean <- heat_clean %>% 
  # standardized average (z-score) for comparing accross occupations. SHows relative exposure and if something is above/below average
  mutate(across(
    c(indoors_controlled, outdoors_exposed, outdoors_undercover),
    ~ as.numeric(scale(.x)),
    .names = "{.col}_scaled"
  )) %>% 
  mutate(total_exposure_z = rowMeans(across(
    c(indoors_controlled_scaled, outdoors_exposed_scaled, outdoors_undercover_scaled)
  ), na.rm = TRUE)) 
```

### Subset Indices

```{r indices}

indices <- heat_clean %>% 
  select(
    industry_occupation,
    average_exposure = total_exposure,
    weighted_exposure,
    total_exposure_z
  )

indices %>% 
  select(
    industry_occupation, total_exposure_z
  ) %>% 
  arrange(desc(total_exposure_z)) %>% 
    kbl(caption = "Standardized Average Exposure Score By Industry and Occupation ACS Category") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) 
```

## Study Area Indices

### Join `heat_clean` to `emp_labor`

```{r indices_join}

emp_labor <- emp_labor %>% 
  left_join(heat_clean, by = "industry_occupation")  %>% 
  relocate(geometry, .after = last_col()) %>% 
  select(-code_2_digit)

```

### Adding Binned Variable

Creating binned categories based on the ONET classification system for each of the three types of heat exposure: indoor with uncontrolled temperatures, outdoor full exposure, and outdoor undercover.

```{r binned_var}

emp_labor <- emp_labor %>% 
mutate(average_score_severity = case_when(
  total_exposure >= 75 ~ "Once a Week or More but Not Every Day",
  total_exposure >= 50 ~ "Once a Month or More but Not Every Week",
  total_exposure >= 25 ~ "Once a Year or More but Not Every Month",
  total_exposure < 25 ~ "Never Exposed",
  TRUE ~ NA_character_
)) %>% 
mutate(weighted_score_severity = case_when(
  weighted_exposure >= 75 ~ "Once a Week or More but Not Every Day",
  weighted_exposure >= 50 ~ "Once a Month or More but Not Every Week",
  weighted_exposure >= 25 ~ "Once a Year or More but Not Every Month",
  weighted_exposure < 25 ~ "Never Exposed",
  TRUE ~ NA_character_
)) %>%
mutate(z_score_severity = case_when(
  total_exposure_z >= 1 ~ "Extreme Exposure",
  total_exposure_z >= 0.5 ~ "High Exposure",
  total_exposure_z >= 0.25 ~ "Moderate Exposure",
  total_exposure_z > 0 ~ "Average Exposure",
  
  total_exposure_z <= -1 ~ "Unlikely to be Exposed",
  total_exposure_z <= -0.5 ~ "Little to No Exposure",
  total_exposure_z <= -0.25 ~ "Limited Exposure",
  total_exposure_z < 0 ~ "Below Average Exposure",
  
  TRUE ~ NA_character_
)) %>%
relocate(geometry, .after = last_col())


```

### Study Area Exposure Summary

Summarize exposure severity across the study area by overall industry exposure and overall occupation exposure

```{r}
exposure_summary <- emp_labor %>%
  group_by(indicator, z_score_severity) %>%
  summarise(employment = sum(ind_occ_employment), .groups = "drop") %>%
  mutate(share = employment / sum(employment), .before = geometry)
```


Repeat the process for tract-level exposure.

## Tract-Level Indices

### Join `heat_clean` to `emp_labor_tract`

```{r tract_heat_join}

emp_labor_tract <- emp_labor_tract %>% 
  left_join(heat_clean, by = "industry_occupation")  %>% 
  relocate(geometry, .after = last_col()) %>% 
  select(-code_2_digit)

```

### Tract Binned Variable

```{r tract_bins}
emp_labor_tract <- emp_labor_tract %>% 
mutate(average_score_severity = case_when(
  total_exposure >= 75 ~ "Once a Week or More but Not Every Day",
  total_exposure >= 50 ~ "Once a Month or More but Not Every Week",
  total_exposure >= 25 ~ "Once a Year or More but Not Every Month",
  total_exposure < 25 ~ "Never Exposed",
  TRUE ~ NA_character_
)) %>% 
mutate(weighted_score_severity = case_when(
  weighted_exposure >= 75 ~ "Once a Week or More but Not Every Day",
  weighted_exposure >= 50 ~ "Once a Month or More but Not Every Week",
  weighted_exposure >= 25 ~ "Once a Year or More but Not Every Month",
  weighted_exposure < 25 ~ "Never Exposed",
  TRUE ~ NA_character_
)) %>%
mutate(z_score_severity = case_when(
  total_exposure_z >= 1 ~ "Extreme Exposure",
  total_exposure_z >= 0.5 ~ "High Exposure",
  total_exposure_z >= 0.25 ~ "Moderate Exposure",
  total_exposure_z > 0 ~ "Average Exposure",
  
  total_exposure_z <= -1 ~ "Unlikely to be Exposed",
  total_exposure_z <= -0.5 ~ "Little to No Exposure",
  total_exposure_z <= -0.25 ~ "Limited Exposure",
  total_exposure_z < 0 ~ "Below Average Exposure",
  
  TRUE ~ NA_character_
)) %>%
relocate(geometry, .after = last_col())

```

### Tract-Level Exposure Share

Computing the total share of employees in each census tract by the severity of standardized exposure score. This score:
1. Collapses exposure employment counts for each `z_score_severity` by taking them from the industry level into the tract level `exposure_emp` with the first summarize call
2. Cleanly recomputes the total exposure denominator grouping by `tract_name` and them dividing by the tract sum employment `sum(exposure_emp)`

It gives us the share of employed workers exposed by level of severity *per tract*. 

```{r tract_exposure_dist}
tract_exposure_dist <- emp_labor_tract %>%
  filter(!is.na(z_score_severity)) %>% 
  group_by(tract_name, z_score_severity) %>%
  summarise(
    exposure_emp = sum(ind_occ_employment, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(tract_name) %>%
  mutate(
    exposure_share = exposure_emp / sum(exposure_emp, na.rm = TRUE)
  ) %>% 
  relocate(geometry, .after = last_col())
```



# References

[@laskaris2025]
